Penalized linear mixed models for structured genetic data

### 2021-05-15 main.tex
updating wording on lasso assumptions re Kai Wang comments

Lasso models, however, rely on an assumption that subjects are independent of each other. In the presence of population structure or cryptic relatedness this assumption is violated: traits are more likely to be similar within a family or ethnic group. The presence of structure can adversely affect analyses in several ways, but the most important concern is that it is likely to introduce false positives. Allele frequency differences due to ancestry within a sample are likely to be mistaken for biological signal if steps are not taken to mitigate this effect. One common solution, dividing a sample and performing ancestry-specific analyses, often results in a loss of statistical power from reduced sample sizes. Furthermore, ancestry-specific analyses cannot easily be applied in the presence of admixture.

@@

As an example, consider a genome wide association study (GWAS) to assess genetic variants associated with lung cancer in a sample comprised of subjects from two distinct subpopulations, A and B. Assume the minor allele of SNP $X$ is more common in subpopulation B compared to subpopulation A, but has no direct effect on lung cancer. Also suppose these subpopulations are geographically segregated in a such a way that subpopulation A is exposed to good air quality, and subpopulation B to poor air quality, and that air quality does have a direct effect on lung cancer. A GWAS of data from these subpopulations would likely find SNP $X$ to be significantly associated with lung cancer. This spurious association is does not arise from the presence of population stratification itself, but because that stratification is related to a causal environmental exposure. Indeed, if subpopulations A and B were not subject to different air qualities, all else being equal, SNP $X$ would not be associated with the phenotype.

@@

A different approach is to use create surrogate variables to represent population structure, and to adjust for these as additional model covariates. Principal component-adjustment (PC-adjustment) methods use the first several eigenvectors a pairwise similarity matrix as these surrogate variables \citep{price2006principal}. This approach may be used in univariate or multivariate frameworks, and allows for analysis on the entirety of a structured sample.

@@

The lasso penalty accomplishes two key things: (1) shrinkage, meaning that coefficients are encouraged to be smaller and less likely to be overestimated and (2) sparsity, meaning that many coefficients will be exactly zero and thus eliminated from the model. Both of these components are key to the lasso's ability to work with large, genome-scale data.  For more information on the lasso, the textbook by \citet{Hastie2009} is a useful reference.  The following sections discuss how to lasso can be modified to account for correlation among individuals.

@@

BIC has been used by several authors for variable selection in penalized models with correlated outcomes \citep{bondell2010joint, schelldorfer2011estimation}. BIC performs poorly in the high-dimensional setting, although modified high-dimensional variants have been proposed \citep{fan2013tuning,Chen2008a}. Cross-validation is also reasonable, with some caveats about the generalizability of prediction error \citep{Rabinowicz2020}. In the present context, this means that although cross-validation is a valid way to estimate prediction error of another sample from the same population, we cannot expect for example cross-validation of a GWAS from European samples to estimate prediction error accurately for calculating genetic risk in Chinese individuals.  Selecting $\lam$ on the basis of FDR is obviously attractive, although there has been little work extending these approaches to correlated outcomes.  Knockoff filters may be useful for this purpose \citep{Sesia2018}, although they have not been studied in the specific context of penalized LMMs.

>>>>

Lasso models, however, are based on a loss function motivated by independence among subjects. In the presence of population structure or cryptic relatedness this is not the case: traits are more likely to be similar within a family or ethnic group. The presence of structure can adversely affect analyses in several ways, but the most important concern is that it is likely to introduce false positives. Allele frequency differences due to ancestry within a sample are likely to be mistaken for biological signal if steps are not taken to mitigate this effect. One common solution, dividing a sample and performing ancestry-specific analyses, often results in a loss of statistical power from reduced sample sizes. Furthermore, ancestry-specific analyses cannot easily be applied in the presence of admixture.

@@

As an example, consider a genome wide association study (GWAS) to assess genetic variants associated with lung cancer in a sample comprised of subjects from two distinct subpopulations, A and B. Assume the minor allele of SNP $X$ is more common in subpopulation B compared to subpopulation A, but has no direct effect on lung cancer. Also suppose these subpopulations are sociodemographically or geographically segregated in a such a way that subpopulation A is exposed to good air quality, and subpopulation B to poor air quality, and that air quality does have a direct effect on lung cancer. A GWAS of data from these subpopulations would likely find SNP $X$ to be significantly associated with lung cancer. This spurious association does not arise from the presence of population stratification itself, but because that stratification is related to a causal environmental exposure. Indeed, if subpopulations A and B were not subject to different air qualities, all else being equal, SNP $X$ would not be associated with the phenotype.

@@

A different approach is to create surrogate variables to represent population structure, and to adjust for these as additional model covariates. Principal component-adjustment (PC-adjustment) methods use the first several eigenvectors a pairwise similarity matrix as these surrogate variables \citep{price2006principal}. This approach may be used in univariate or multivariate frameworks, and allows for analysis on the entirety of a structured sample.

@@

The lasso penalty accomplishes two key things: (1) shrinkage, meaning that coefficients are encouraged to be smaller and less likely to be overestimated and (2) sparsity, meaning that many coefficients will be exactly zero and thus eliminated from the model. Both of these components are key to the lasso's ability to work with large, genome-scale data.  For more information on the lasso, the textbook by \citet{Hastie2009} is a useful reference.  The following sections discuss how the lasso can be modified to account for correlation among individuals.

@@

BIC has been used by several authors for variable selection in penalized models with correlated outcomes \citep{bondell2010joint, schelldorfer2011estimation}. BIC performs poorly in the high-dimensional setting, although modified high-dimensional variants have been proposed \citep{fan2013tuning, chen2008}. Cross-validation is also reasonable, with some caveats about the generalizability of prediction error \citep{Rabinowicz2020}. In the present context, this means that although cross-validation is a valid way to estimate prediction error of another sample from the same population, we cannot expect for example cross-validation of a GWAS from individuals of European ancestry to estimate prediction error accurately for calculating genetic risk in individuals of Asian ancestry.  Selecting $\lam$ on the basis of FDR is obviously attractive, although there has been little work extending these approaches to correlated outcomes.  Knockoff filters may be useful for this purpose \citep{Sesia2018}, although they have not been studied in the specific context of penalized LMMs.
